---
title: "elNamed() to safely subset from named lists and vectors"
date: "2025-06-19"
knit: litedown:::knit
---

<!-- pre-hook -->
```{r pre-hook, child = "_prehook.Rmd", include = FALSE, order = 0}
```

I've always been annoyed by how subsetting missing elements by name had different behaviors for lists vs. vectors.

## Base behaviors

For named lists, subsetting with a non-existent name returns a `NULL`.

```{r}
list(a = 1)[["b"]]
```

For named vectors, it errors with `[[`.

```{r, error = TRUE}
c(a = 1)[["b"]]
```

The case of `[` is interesting because it returns a `NA`.

```{r}
c(a = 1)["b"]
```

I get why, but it's not so useful. It breaks nice equivalencies between lists and vectors.

## `elNamed()` to subset by name

Then I discovered `elNamed()` from `{methods}`. It's a sister of `el()` that subsets by name (vs. position):

```{r}
el(c(a = 1), 1)

elNamed(c(a = 1), "a")
```

What's kinda cool about `elNamed()` is that it returns `NULL` for unmatched names, for _both_ lists and vectors.

```{r}
elNamed(list(a = 1), "b")

elNamed(c(a = 1), "b")
```

And you can even turn it into an assertion!

```{r, error = TRUE}
elNamed(list(a = 1), "b", mustFind = TRUE)

elNamed(c(a = 1), "b", mustFind = TRUE)
```

::: {.side .side-right}
`?elNamed` says it's for "compatibility with S+"!
:::

I don't actually use `elNamed()` in practice, though.

But cool find nonetheless :)

---

This post was adapted from a Mastodon exchange: https://fosstodon.org/@yjunechoe/111383917758940010

<!-- post-hook -->
```{r post-hook, child = "_posthook.Rmd"}
```
