---
title: "Recover the position of elements in a ggrepel layer"
date: "2025-06-08"
knit: litedown:::knit
---

<!-- pre-hook -->
```{r pre-hook, child = "_prehook.Rmd", include = FALSE, order = 0}
```

My take at a frequently asked question: "How can you recover the positions calculated by a [ggrepel](https://ggrepel.slowkow.com/) layer?"

# Setup

```{r}
library(ggplot2)
library(ggrepel)
p <- ggplot(mtcars,
            aes(wt, mpg, label = rownames(mtcars), colour = factor(cyl))) +
  geom_point() +
  geom_text_repel(seed = 1)
p
```


# When are the positions calculated?

Not available at `ggplot_build()`. At this point, `x` and `y` are the un-transformed data values, identical to the original data columns:

```{r, collapse = TRUE}
repel_layer_data <- layer_data(p, 2L)
all.equal(
  mtcars[, c("wt", "mpg")],
  repel_layer_data[, c("x", "y")],
  check.attributes = FALSE,
  use.names = FALSE
)
```

::: {.side .side-right}
A layer's grob (à la `layer_grob()`) is a list of grobs, one for each panel. That's why we're working with a length-1 list here.
:::

Also not available at `ggplot_gtable()`, where a placeholder grob exists for the layer but no labels have been materialized yet:

```{r, collapse = TRUE}
repel_layer_grob <- layer_grob(p, 2L)
repel_layer_grob[[1]]
repel_layer_grob[[1]]$children
```

Calculated positions ARE available upon rendering, at `grid.draw()`. The elements drawn onto the current device can be inspected using `grid.ls()`.

::: {.side .side-right}
For ggplots, a prior call to `grid.force()` is necessary to make visible graphical objects which are computed at draw time.
:::

```{r, fig.path = NA}
library(grid)
plot(p) # Plot suppressed here for space
grid.force()
grid.ls()
```

I use the term "computed at draw time" to loosely refer to the work of `grid::makeContext()` and `grid::makeContext()` methods. These allow you to lazily implement **adaptive** logic using information from the current device, like "Draw a circle of 1 inch radius, and color it red if it exceeds the size of the device."

In the context of ggrepel, the positioning work is done inside `ggrepel:::makeContent.textrepeltree()` which also populates layer grob, of class `textrepeltree`

```{r, collapse = TRUE, fig.path = NA}
repel_grob <- grid.get("textrepeltree", grep = TRUE)
class(repel_grob)
repel_grob
```


# What are the position values?

The `textrepeltree` grob which represents the repel layer's grob is a `gTree`, meaning a collection of grobs. In our example, there are segments (`segmentrepelgrob`) and text (`textrepelgrob`) elements involved in the layer:

```{r}
repel_grob$children
```

Since these are the actually rendered grobs, they have position values associated with them. We can grab one of the text grobs to inspect it:

```{r, collapse = TRUE}
repel_grob_text1 <- getGrob(repel_grob, "text", grep = TRUE)
repel_grob_text1$x
repel_grob_text1$y
```

The grob also includes the label and color, which we will need for reconstruction later:

```{r, collapse = TRUE}
repel_grob_text1$label
repel_grob_text1$gp$col
```

Lastly, note also that ggrepel attaches additional context from `ggplot_build()` to the `$data` element of the repel layer's grob. But use this with care: the `x` and `y` column values **are not accurate here** since repel is only calculated at render time.

```{r, print.args = list(data.frame = list(limit = c(10, 6)))}
repel_grob$data
```


# Reconstructing repelled text data

We first write a helper to extract relevant features from a text grob:

```{r, fig.path = NA}
process_textGrob <- function(g) {
  data.frame(
    x = convertUnit(g$x, "native", valueOnly = TRUE),
    y = convertUnit(g$y, "native", valueOnly = TRUE),
    label = g$label,
    colour = g$gp$col
  )
}
process_textGrob(repel_grob_text1)
```

Then apply this over text grobs in the repel layer:

```{r, fig.path = NA}
repel_text_df <- repel_grob$children |>
  Filter(x = _, \(g) grepl(x = g$name, "textrepelgrob")) |>
  lapply(process_textGrob) |>
  do.call(rbind, args = _)
repel_text_df
```


# Re-plotting repelled text

::: {.side .side-right}
This "same dimension" constraint is actually not entirely accurate - you could technically convert npc coordinates back into data space by back-transforming with plot scales in `p$layout$panel_scales_(x|y)`. But that's a more brittle and involved process.
:::

It should go without saying that the repelled text can only be replicated in a new plot if it has the **same dimensions** as the original plot where the text positions were recovered from.

As long as that is satisfied, we can replicate the original repel layer (sans the segments, since we haven't dealt with that).

Take care to use `I()`, especially for `x` and `y`, so that the scales leave the values un-transformed.

```{r}
p2 <- ggplot(mtcars,
            aes(wt, mpg, label = rownames(mtcars), colour = factor(cyl))) +
  geom_point() +
  geom_text(
    aes(x = I(x), y = I(y), label = label, color = I(colour)),
    data = repel_text_df
  )
p2
```

<!-- post-hook -->
```{r post-hook, child = "_posthook.Rmd"}
```
